---
title: "Week 3 Problems"
author:
  - name: 
      given: Juan Pablo
      family: Fonseca-Zamora
    id: jpfz
    orcid: 0009-0006-7042-3828
    email: ts18jpf@leeds.ac.uk
    affiliation: 
      - name: University of Leeds
format: 
  pdf
pdf-engine: xelatex
mainfont: Nunito Sans Light 
sansfont: Nunito Sans
monofont: Fira Code
papersize: a4
lang: en-GB
geometry:
  - top=30mm
  - left=25mm
  - right=25mm
  - bottom=25mm
---
```{r,message=FALSE,include=FALSE}
library(rethinking)
library(dagitty)
```

**1.** The first two problems are based on the same data. The data in data(foxes)
are 116 foxes from 30 different urban groups in England. These fox groups
are like street gangs. Group size (groupsize) varies from 2 to 8 individuals.
Each group maintains its own (almost exclusive) urban territory. Some territories
are larger than others. The area variable encodes this information.
Some territories also have more avgfood than others. And food influences
the weight of each fox. Assume this DAG:

```{r, echo=FALSE,fig.dim=c(1.5,1.5)}
dag_foxes <- dagitty("dag{A->F;F->G;F->W;G->W }")

coordinates(dag_foxes) <- list(x=c(A=1,W=1,F=0,G=2),y=c(A=0,W=2,F=1,G=1))

drawdag(dag_foxes)
```
where $F$ is avgfood, $G$ is groupsize, $A$ is area, and $W$ is weight.

Use the backdoor criterion and estimate the total causal influence of $A$ on
$F$. What effect would increasing the area of a territory have on the amount
of food inside it?

***Solution***
Based on the DAG, there are not other variables affecting A or mediating the effect
of A on F, therefore it would be enough to produce a linear model

```{r}
data("foxes")
d <- foxes

d$G <- scale(d$groupsize)
d$A <- scale(d$area)
d$W <- scale(d$weight)
d$F <- scale(d$avgfood)

m1_a <- quap(
  alist(
    F ~ dnorm(mu,sigma),
    mu  <- a + b_A*A,
    a ~ dnorm(0,1),
    b_A ~ dnorm(0,2),
    sigma ~ dexp(1)
    ),
  data = d
)

precis(m1_a)
```
Based on the results, there is a positive effect of the change in area on the food availability, i.e. an increase of area is linked to an increase of avg food.

**2.** Infer the total causal effect of adding food F to a territory on the weight
W of foxes. Can you calculate the causal effect by simulating an intervention
on food?

***Solution***
There are two paths through which food availability affects the weight of foxes. To determine
the total causal effect it would be enough just to use $F$ and $W$

```{r}
m2 <- quap(
  alist(
    W ~ dnorm(mu,sigma),
    mu  <- a + b_F*F,
    a ~ dnorm(0,1),
    b_F ~ dnorm(0,2),
    sigma ~ dexp(1)
    ),
  data = d
)

precis(m2)
```
The total casual effect of $F$ is not conclusive.


**3.** Infer the direct causal effect of adding food F to a territory on the weight
W of foxes. In light of your estimates from this problem and the previous
one, what do you think is going on with these foxes?

***Solution***
In this case, there is a *pipe* formed by F->G->W; so, to estimate the direct casual effect
it is necessary to stratify by $G$

```{r}
m3 <- quap(
  alist(
    W ~ dnorm(mu,sigma),
    mu  <- a + b_F*F + b_G*G,
    a ~ dnorm(0,1),
    b_F ~ dnorm(0,2),
    b_G ~ dnorm(0,2),
    sigma ~ dexp(1)
    ),
  data = d
)

precis(m3)
```

The direct effect of $F$ on $W$ is positive, while the effect of $G$ on $W$ is negative, which explains the results of the previous model.
